---
title: "Zambia Comparisons"
author: "Danny Parsons"
date: "11/06/2020"
output: 
  html_document:
    fig_width: 12
    fig_height: 6
---

```{r packages, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(here)
library(ggplot2)
library(lubridate)
library(reshape2)
library(viridis)
library(RColorBrewer)
library(tidyr)
library(hydroGOF)
library(stringr)
library(knitr)
library(kableExtra)
library(rnaturalearthdata)
library(rnaturalearth)
library(ggrepel)
library(sp)
library(tibble)
library(verification)
library(purrr)
library(dplyr)
```

```{r setup, include=FALSE}
source(here("src", "helper_funs.R"))

zm_rds <- readRDS(here("data", "satellite", "merged_data.RDS"))
zm <-zm_rds$get_data_frame(data_name = "final_merged") 
if(anyDuplicated(zm %>% dplyr::select(station, date))) stop("Duplicates found!")

zm <- zm %>% 
  select(station, date, rain, chirps_rain, chirp_rain, era5_rain, tamsat_rain)
# 1 Aug = 214
s_doy_start <- 214
zm <- zm %>% mutate(doy = yday_366(date),
                    year = year(date),
                    month = month(date),
                    s_doy = (doy - s_doy_start + 1) %% 366,
                    s_doy = ifelse(s_doy == 0, 366, s_doy),
                    syear = year,
                    syear = ifelse(s_doy > (366 - s_doy_start + 1), syear - 1, syear),
                    month = factor(month, levels = c(8:12, 1:7)),
                    month_abb = factor(month, labels = month.abb[c(8:12, 1:7)]),
                    rain = ifelse(rain < 0, 0, rain),
                    era5_rain = ifelse(era5_rain < 0, 0, era5_rain),
                    chirps_rain = ifelse(chirps_rain < 0, 0, chirps_rain), 
                    chirp_rain = ifelse(chirp_rain < 0, 0, chirp_rain), 
                    tamsat_rain = ifelse(tamsat_rain < 0, 0, tamsat_rain)
                    )

zm_long_st <- zm %>% 
  melt(id.vars = c("station", "date", "year", "syear", "month", "month_abb", 
                   "doy", "s_doy", "rain"),
       measure.vars = names(zm)[endsWith(names(zm), "rain")][-1],
       variable.name = "product", value.name = "pr_rain")

zm_long <- zm %>% 
  melt(id.vars = c("station", "date", "year", "syear", "month", "month_abb", "doy", "s_doy"),
       measure.vars = names(zm)[endsWith(names(zm), "rain")],
       variable.name = "product", value.name = "rain") %>%
  mutate(rainday = rain > 1)

zm_long$product <- recode(zm_long$product, rain = "station")
zm_long$station <- recode(zm_long$station, CHIPAT01 = "Chipata", PETAUK01 = "Petauke")

stations <- c("Moorings", "Choma", "Kasama", "Livingstone", "Magoye", "Mansa", "Mpika", "Chipata", "Petauke")
products <- levels(zm_long$product)
products <- products[-1]
names(products) <- substr(products, 1, nchar(products) - 5)

metadata_zm <- readRDS(here("data", "station", "zambia_metadata.RDS"))
metadata_zm$station <- recode(metadata_zm$station, CHIPAT01 = "Chipata", PETAUK01 = "Petauke")

metadata_zm$station <- factor(metadata_zm$station, levels = stations)
zm_long$station <- factor(zm_long$station, levels = stations)

by_station <- zm_long %>%
  group_by(station) %>%
  filter(!is.na(rain) & product == "station") %>%
  summarise(first_date = first(date),
            last_date = last(date))

metadata_zm <- left_join(metadata_zm, by_station, by = "station")
rm(by_station)

zm_long <- left_join(zm_long, metadata_zm, by = "station")

zm_long <- zm_long %>% 
  filter(date >= first_date & date <= last_date)

skable <- function(kable_input) {
  kable_input %>% kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                                full_width = FALSE)
}
```

## Stations

- Mansa and Kasama are the only station in the North. 
- It's results are different to the other three, which could be the location or the quality of the station
- Would be useful to have further stations in the North
- Station data filtered to start in 1979 (earliest product year)

```{r}
metadata_zm %>% 
  dplyr::select(station, latitude, longitude, first_date, last_date) %>%
  kable(digits = 2) %>%
  skable()
```


```{r station_map, fig.width=9, fig.height=4.5}
sf_zm <- ne_countries(country = "Zambia", returnclass = "sf")
ggplot(sf_zm) + 
  geom_sf() +
  geom_point(data = metadata_zm, aes(x = longitude, y = latitude)) +
  geom_text_repel(data = metadata_zm, aes(x = longitude, y = latitude, label = station))
rm(sf_zm)
```

```{r gof_fun}
dgof <- function(df, sim, obs, na.rm = TRUE) {
  g <- hydroGOF::gof(sim = df[[sim]], obs = df[[obs]], na.rm = na.rm)
  as.list(g[ ,1])
}

comp_stats <- c("r", "ME", "PBIAS %", "MAE", "NSE", "rSD")
names(comp_stats) <- c("Correlation coefficient (1 = Perfect)",
                       "Mean bias (same units)",
                       "Percentage bias (%)",
                       "Mean absolute bias (same units)",
                       "Nash-Sutcliffe efficiency (1 = Perfect)",
                       "Ratio of standard deviations (< 1 less variable, > 1 more variable)")

comp_stats_digits <- c(2, 0, 0, 0, 2, 3)
```

## Yearly Comparisons (August to July)

```{r yearly_calcs}
by_syear <- zm_long %>%
  group_by(station, syear, product) %>%
  summarise(total_rain = sum(naif_nmin(rain, 355)), 
            n_rain = sum(naif_nmin(rain, 355) > 0.1), 
            max_rain = max(naif_nmin(rain, 350)), 
            mean_rain = total_rain/n_rain, 
            n_na = sum(is.na(rain))
            )

by_syear_st <- by_syear %>%
  pivot_wider(id_cols = c(station, syear), 
              names_from = product, values_from = total_rain:n_na, names_sep = "__") %>%
  pivot_longer(cols = -c(station, syear, ends_with("station")), 
               names_to = c(".value", "product"), names_sep = "__")

gof_syear <- by_syear_st %>%
  group_by(station, product) %>%
  nest() %>%
  mutate(n = purrr::map_int(data, 
                        ~sum(!is.na(.$total_rain__station) & !is.na(.$total_rain))),
         gof__total_rain = purrr::map(data, dgof, "total_rain", "total_rain__station", 
                                      na.rm = TRUE),
         gof__n_rain = purrr::map(data, dgof, "n_rain", "n_rain__station", na.rm = TRUE),
         gof__mean_rain = purrr::map(data, dgof, "mean_rain", "mean_rain__station", 
                                     na.rm = TRUE)
         )

gof_pr <- gof_syear %>% 
  unnest(cols = data) %>% 
  group_by(product) %>% 
  nest()
```

```{r yearly_plots_fun}
yearly_plots <- function(df, gof_col, stat_pr, stat_st, product_name) {
  max_y <- max(c(df[[stat_pr]], df[[stat_st]]), na.rm = TRUE)
  vals_relace <- c(product_name, "station")
  names(vals_relace) <- c(stat_pr, stat_st)
  dat <- df %>% 
    pivot_longer(cols = c(stat_pr, stat_st), names_to = "product", values_to = stat_pr) %>%
    mutate(product = recode(product, !!!vals_relace),
           ME = purrr::map_dbl(.data[[gof_col]], "ME"),
           r = purrr::map_dbl(.data[[gof_col]], "r"),
           rSD = purrr::map_dbl(.data[[gof_col]], "rSD")
           )
  mean_df <- dat %>% 
    group_by(station, product) %>% 
    summarise(m = mean(.data[[stat_pr]], na.rm = TRUE))
  g <- ggplot(dat, aes(x = syear, y = .data[[stat_pr]], colour = product)) +
    geom_line() +
    geom_point() +
    geom_hline(data = mean_df, aes(yintercept = m, colour = product)) +
    scale_x_continuous(limits = c(1979, 2012)) +
    # n
    geom_text(data = dat, aes(label = paste("n", n)), size = 4,
              x = 1979, y = max_y, na.rm = TRUE, 
              inherit.aes = FALSE) +
    # bias
    geom_text(data = dat, aes(label = paste("bias", signif(ME, 2))), 
              size = 4, x = 1979 + 6, y = max_y, na.rm = TRUE, 
              inherit.aes = FALSE) +
    # cor
    geom_text(aes(label = paste("cor", round(r, 2))), 
              size = 4, x = 1979 + 14, y = max_y, na.rm = TRUE, 
              inherit.aes = FALSE) +
    # rSD
    geom_text(aes(label = paste("rSD", round(rSD, 2))), 
              size = 4, x = 1979 + 22, y = max_y, na.rm = TRUE, 
              inherit.aes = FALSE) +
    ggtitle(paste(stat_pr, ":", product_name, "vs", "station")) +
    facet_wrap(~station)
  g
}
```

```{r stats_tables_fun}
stats_tables <- function(df, obj_col, obj_stats = comp_stats) {
  for(i in seq_along(obj_stats)) {
    dat <- df %>% 
      ungroup() %>%
      mutate(purrr::map_dbl(df[[obj_col]], obj_stats[i]))
    names(dat)[ncol(dat)] <- obj_stats[i]
    dat <- dat %>% 
      mutate(station = as.character(station)) %>%
      pivot_wider(id_cols = "station", names_from = "product", 
                  values_from = obj_stats[[i]]) %>%
      data.frame()
  dat[nrow(dat) + 1, ] <- c(list("(All)"), as.list(as.numeric(colMeans(dat[ , -1]))))
  dat %>%
    kable(digits = comp_stats_digits[i], caption = names(obj_stats[i]), 
          format.args = list(big.mark = ",")) %>%
    skable() %>%
    row_spec(nrow(dat), bold = TRUE) %>%
    print()
  }
}
```

### Comparison Statistics for Total Yearly Rainfall

- In general all products follow a similar shape each year (high correlation)
- All products, apart from ERA5, estimate the mean quite well (low bias and low percentage bias).
- ERA5 overestimates (high bias) but could be expected as it represents the largest area.
- RFE2 and IMERG look promising across all measures, but few years compared, need the extended station records for better comparison
- Aside from RFE2 and IMERG, CHIRPS performs well over all measures
- Slightly less variability in all products, between 70% - 90% (apart from RFE2 which has roughly the same variability)
- Results for Mansa are generally different across products, indicating possible issues with the station records

```{r yearly_total_plots}
p_syear_total <- gof_pr %>%
  mutate(p = purrr::map(data, yearly_plots, gof_col = "gof__total_rain", 
                        stat_pr = "total_rain", stat_st = "total_rain__station", 
                        product_name = product),
         paths = here("results", "zambia", 
                      paste0("zambia_", "syear_", "total_rain", "_", product, ".png")))

walk2(p_syear_total$paths, p_syear_total$p, ggsave, width = 12, height = 6)
walk(p_syear_total$p, print)
```

```{r yearly_n_obs, results="asis"}
dat <- gof_syear %>% 
  dplyr::select(station, product, n) %>%
  pivot_wider(id_cols = c(station), names_from = product, values_from = n)

names(dat)[endsWith(names(dat), "rain")] <- 
  substr(names(dat)[endsWith(names(dat), "rain")], 
         1, nchar(names(dat)[endsWith(names(dat), "rain")]) - 5)
dat %>%
  kable(caption = "Number of years compared") %>%
  skable()
```

```{r yearly_total_rain_tables, results="asis"}
stats_tables(gof_syear, "gof__total_rain")
```


### Comparison Statistics for Number of rainy days

- Reasonably good correlation across the products, so similar pattern each year
- All products overestimate number of rainy days with the same threshold
- Wide variation of the bias amount across products, ERA5 has largest bias, CHIPS has lowest
- Almost the same variability for all products
- Results for Mansa are generally different across products, indicating possible issues with the station records

```{r yearly_n_rain_plots}
p_syear_n <- gof_pr %>%
  mutate(p = purrr::map(data, yearly_plots, gof_col = "gof__n_rain", 
                        stat_pr = "n_rain", stat_st = "n_rain__station", 
                        product_name = product),
         paths = here("results", "zambia", 
                      paste0("zambia_", "syear_", "n_rain", "_", product, ".png")))

walk2(p_syear_n$paths, p_syear_n$p, ggsave, width = 12, height = 6)
walk(p_syear_n$p, print)
```

```{r yearly_n_rain_tables, results="asis"}
stats_tables(gof_syear, "gof__n_rain")
```

## Monthly Comparisons (November to April)

```{r monthly_calcs}
by_month <- zm_long %>%
  group_by(station, syear, month_abb, product) %>%
  filter(month %in% c(11:12, 1:4)) %>%
  summarise(total_rain = sum(naif_nmin(rain, 25)), 
            n_rain = sum(naif_nmin(rain, 25) > 0.1), 
            max_rain = max(naif_nmin(rain, 20)), 
            mean_rain = total_rain/n_rain, 
            n_na = sum(is.na(rain))
            )

by_month_st <- by_month %>%
  pivot_wider(id_cols = c(station, syear, month_abb), 
              names_from = product, values_from = total_rain:n_na, names_sep = "__") %>%
  pivot_longer(cols = -c(station, syear, month_abb, ends_with("station")), 
               names_to = c(".value", "product"), names_sep = "__")

gof_month <- by_month_st %>%
  group_by(station, product) %>%
  nest() %>%
  mutate(n = purrr::map_int(data, 
                        ~sum(!is.na(.$total_rain__station) & !is.na(.$total_rain))),
         gof__total_rain = purrr::map(data, dgof, "total_rain", "total_rain__station", 
                                      na.rm = TRUE),
         gof__n_rain = purrr::map(data, dgof, "n_rain", "n_rain__station", na.rm = TRUE),
         # gof__mean_rain = purrr::map(data, dgof, "mean_rain", "mean_rain__station", 
         #                             na.rm = TRUE)
         )

gof_pr_month <- gof_month %>%
  unnest(cols = data) %>% 
  group_by(product) %>% 
  nest()
```

```{r monthly_plots_fun}
monthly_plots <- function(df, stat_pr, stat_st, product_name) {
  vals_relace <- c(product_name, "station")
  names(vals_relace) <- c(stat_pr, stat_st)
  dat <- df %>% 
    filter(month_abb %in% month.abb[c(11:12, 1:4)]) %>%
    pivot_longer(cols = c(stat_pr, stat_st), names_to = "product", values_to = stat_pr) %>%
    mutate(product = recode(product, !!!vals_relace))

  mean_df <- dat %>% 
    group_by(station, product, month_abb) %>% 
    summarise(m = mean(.data[[stat_pr]], na.rm = TRUE))

  g <- ggplot(dat, aes(x = syear, y = .data[[stat_pr]], colour = product)) +
    geom_line() +
    geom_point() +
    #geom_hline(data = mean_df, aes(yintercept = m, colour = product)) +
    scale_x_continuous(limits = c(1979, 2012)) +
    ggtitle(paste(stat_pr, ":", product_name, "vs", "station")) +
    facet_grid(station~month_abb)
  g
}
```

### Comparison statistics for total monthly rainfall
- In general all products follow a similar shape each year (high correlation)
- All products, apart from ERA5, estimate the mean quite well (low bias and low percentage bias), with a very slight underestimation compared to yearly totals
- ERA5 overestimates (high bias) but could be expected as it represents the largest area.
- Aside from RFE2 and IMERG, CHIRPS performs well over all measures
- Slightly less variability in all products, between 80% - 100%
- Generally higher correlations and better s.d. ratios than yearly totals
- Results for Mansa are generally different across products, indicating possible issues with the station records

```{r monthly_plots_total_rain}
p_month_total <- gof_pr_month %>%
  mutate(p = purrr::map(data, monthly_plots, stat_pr = "total_rain", 
                        stat_st = "total_rain__station", product_name = product),
         paths = here("results", "zambia", 
                      paste0("zambia_", "month_", "total_rain", "_", product, ".png")))

walk2(p_month_total$paths, p_month_total$p, ggsave, width = 12, height = 12)
walk(p_month_total$p, print)
```

```{r monthly_tables_total_rain, results="asis"}
stats_tables(gof_month, "gof__total_rain")
```

### Comparison statistics for number of monthly raindays
- Generally high correlation for all products
- As shown in the Markov Chain models, the bias varies across products, between 1 - 14 days per month. ARC2 & CHIRPS lowest (1-2 days), TAMSAT (3 days), IMERG & RFE2 (6-7 days), ERA5 (14 days)
- Generally the products have the same or slightly higher variability

```{r monthly_plots_n_rain}
p_month_total <- gof_pr_month %>%
  mutate(p = purrr::map(data, monthly_plots, stat_pr = "n_rain", 
                        stat_st = "n_rain__station", product_name = product),
         paths = here("results", "zambia", 
                      paste0("zambia_", "month_", "n_rain", "_", product, ".png")))

walk2(p_month_total$paths, p_month_total$p, ggsave, width = 12, height = 12)
walk(p_month_total$p, print)
```

```{r monthly_tables_n_rain, results="asis"}
stats_tables(gof_month, "gof__n_rain")
```